Learning to Synthesize Rules for Network Intrusion Detectors

### 2020-03-03 paper/paper.tex
pass on algo
pass

\begin{figure}[h!]
  \centering
  \includegraphics[trim=0 395 50 0,clip,width=0.48\textwidth]{figs/nids-workflow}
  \caption{The \tname\ workflow.}
  \label{fig:overview}
  \vspace{-2ex}
\end{figure}

@@

\subsection{Step 2: Rule Minimization}
\label{sec:minimization}

\Fix{consider changing the name of this step}
The seed rule produced in the first step satisfies the criterion to
only capture positive traffic. However, that rule is overconstrained.
It could include options that are irrelevant to capturing the attack
and could lead to false negatives---the \nids\ may be unable to
capture manifestations of the same attack whose messages do not
exhibit the data satisfying irrelevant options.

@@

\tname{} leverages public data sets~\cite{tcpreplay} of benign traffic
to guide the search for more general rules, hence addressing the
overffiting issue. To that end, \tname{} \emph{minimizes} the rules
produced in the previous step, preserving the invariant of \emph{only}
capturing the positive malicious traffic. We say that a rule is
minimal if discarding any option results in the capture of negative
(\ie{}, benign) traffic.  The minimization procedure we propose in
this step builds on the observation that all options from the rule
produced in the previous step are satisifed by the positive
traffic. Consequently, any subset of these options also captures the
positive traffic. The minimization procedure that we describe in the
following iteratively discards options from visited rules---derived
from the seed rule---using the negative traffic to guide the search
for minimal rules.

\algdef{SE}[DOWHILE]{Do}{doWhile}{\algorithmicdo}[1]{\algorithmicwhile\ #1}%
\algnewcommand\algorithmicforeach{\textbf{for each}}
\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
\algnewcommand\And{\textbf{and} }%
\setlength{\textfloatsep}{3pt}% Remove \textfloatsep
\begin{algorithm}[t!]
  \noindent\hspace{-25ex}
  \textbf{INPUT:} (overconstrained) seed rule $sr$ \\
  \noindent\hspace{-27.5ex} 
  \textbf{OUTPUT:} Set of minimized rules $\mathcal O$\\
%\vspace{-3ex}
\captionof{algorithm}{Minimization algorithm}\label{algo1}
\begin{algorithmic}[1]
  \State $\mathcal R \gets \{\mathit{sr}\}$\Comment{initial population
    of candidate solutions contains only the input
    rule.}\label{line-init}
  \State $\mathcal O \gets \emptyset$
  \Do\label{start-do-while}
      \State $\mathcal R' \gets \emptyset$ \Comment{new iteration. reset.}\label{init-rprime}
      \ForEach {$ro \in \mathcal R $}\Comment{analyze each candidate rule.}\label{start-foreach}
           \For {$i \gets 1$ to $N$}\Comment{mutate each rule $N$ times.}\label{start-for}
               \State $r \gets copy(ro)$\label{copy}
               \State $j\gets \mathit{rand}(0,\mathit{len}(r))$\label{choose-index}
               \State $r[j]\gets 0$\Comment{discard option $j$ in $r$}\label{delete-option}
               \If{$\mathit{isNew(r)} \And\mathit{isPlausible}(r)$} 
                   \State $\mathcal R' \gets \mathcal R' \cup \{r\}$\label{line-progress}
               \EndIf
           \EndFor\label{end-for}
       \EndFor\label{end-foreach}
       \State $\mathcal R \gets \mathcal R'$\Comment{update new generation}
       \State $\mathcal O \gets \mathcal O \cup \mathcal
       R$\Comment{remember all rules}
  \doWhile{$\mathcal R\not= \{\}$}\Comment{repeat until reaching a fixpoint}\label{end-do-while}
\end{algorithmic}
\end{algorithm}

Algorithm~\ref{algo1} shows the pseudocode for \tname{}'s rule
minimization. The algorithm takes as input an overconstrained rule
produced by the reverse engineering step and produces a set $\mathcal
O$ of minimized rules. Line~\ref{line-init} initializes the current
population of individuals (\ie{}, rules) $\mathcal{R}$ with the seed
rule $sr$. We encode a rule as a bitvector where each index in the
vector represents one option; the value assigned at a given position
indicates whether or not the option is present in the rule. One
iteration of the outer loop
(lines~\ref{start-do-while}-\ref{end-do-while}) processes the current
population of individuals $\mathcal R$ and creates the next generation
$\mathcal R'$. Line~\ref{init-rprime} initializes the population of
individuals for the next generation. The inner loop
(lines~\ref{start-foreach}--\ref{end-foreach}) iterates through each
candidate rule in $\mathcal R$. The algorithm mutates each of these
rules for $N$ times, discarding one option of the rule each
time. Line~\ref{copy} assigns a copy of the bitvector encoding a rule
in $\mathcal R$ to variable $r$. Function $\mathit{len}$ returns the
length of a vector and function $\mathit{rand}$ randomly chooses one
integer value in the range defined by its parameters. So, the
expression $\mathit{rand(0, len(r))}$, appearing at
line~\ref{choose-index}, returns an index of the option in $r$ to
delete at line~\ref{delete-option}. The call to $\mathit{isNew}$
checks if $r$ has not been previously generated through another path
whereas the call to $\mathit{isPlausible}$ checks if the rule only
captures positive traffic. If $r$ satisfies both criteria, it is added
to $\mathcal R'$, which denotes the next generation of individuals in
the search.

The algorithm terminates when no rule in the current
generation of individuals could be further reduced without violating
the property that a candidate solution should only capture positive
traffic. This means that execution of one iteration has not reached
the statement declared at line~\ref{line-progress}, which adds one
element to the set $\mathcal{R'}$. Consequently, as $\mathcal{R'}$ is
initialized empty that set will remain empty at the end of an
iteration of the outer loop.

Note that the algorithm produces no rules in the last iteration. Rules
with a single option are produced in the second-to-last iteration. In
other iterations, ules with more than one option are
produced. Conceptually, the search induces a search tree with the seed
rule $sr$ as the root, single-option rules as child nodes, and rules
with more options as internal nodes. The edges of the tree denote the
ancestor relation between two rules.

It is also worth noting that the algorithm reports every rule in the
tree instead of only those that are surely minimal, \ie{} rules with
one option. The reason is that the the traffic used to check
plausibility is limited by definition---reporting only minimal rules
with respect to that traffic could discard the correct rule. In
summary, this algorithm reports many rules that are plausible by using
the benign traffic to reduce the overconstrained rule produced at the
\reveng\ step. In the following step, \tname{} ranks the rules this
algorithm reports by their similarity to existing trusted
rules. Intuitively, rules that do not look alike to rules written by
humans and previously used rank low.

\begin{proposition}
  \vspace{-1ex}  
  The minimization algorithm reports rules that capture only the
  positive traffic.
\end{proposition}
\begin{proof}
  \vspace{-1.5ex}
  Every rule created during the search captures the positive malicious
  traffic as every rule contains a proper subset of the options from
  the original rule and all those options are satisfied by the
  malicious traffic. By definition of function $\mathit{isPlausible}$,
  these rules also capture \emph{only} the positive traffic. If
  negative traffic is captured, a rule is not added to $\mathcal R'$;
  consequently, it cannot be an element of $\mathcal O$ and reported
  on the output.
  \vspace{-1ex}    
\end{proof}

@@

\textbf{Implementation details.} We implemented function
\emph{isPlausible} by 1)~updating Suricata to only monitor the rule
that is passed as parameter\Mar{isto realmente eh necessario? se sim,
  explique. me pareceu que o que vc. precisa ate aqui jah estaria no
  log do \suri}, 2)~replaying the negative benign traffic (see
Section~\ref{sec:dataset-benign}), and 3)~reading the Suricata logs to
check if any benign message was captured by the rule. It is worth
noting that the benign traffic is stored in \emph{pcap}
files~\cite{pcap} and Suricata can process those files very
efficiently compared to analyzing network messages directly. Instead
of sending traffic to a machine with a deployed server (\eg{}, web
server) that \suri\ would monitor, our setup calls \suri\ directly
through an API. \suri\ is invoked multiple times in the process of
minimizing the rule.  \textbf{Limitations.} It is worth noting that
although \tname{} is accurate for generating rules that discrimate
positive and negative traffic, the quality of these rules is limited
by the quality of the training data it uses. For the positive traffic,
\tname{} only requires one sample, but can benefit of several
samples\Mar{como? nao estah claro. por favor expliquem.}. The same
applies to the negative traffic. However, in this case, \tname{} uses
different public data sets available online (see
Section~\ref{sec:objectanalysis-benign}).

>>>>

\begin{figure}[h!]
  \centering
  \includegraphics[trim=0 395 50 0,clip,width=0.48\textwidth]{figs/nids-workflow}
  \caption{The \tname\ workflow.\Mar{revisar}}
  \label{fig:overview}
  \vspace{-2ex}
\end{figure}

@@

\subsection{Step 2: Rule Creation}
\label{sec:minimization}

The seed rule produced in the first step satisfies the criterion to
only capture positive traffic. However, that rule is overconstrained.
It could include options that are irrelevant to capturing the attack
and could lead to false negatives---the \nids\ may be unable to
capture manifestations of the same attack whose messages do not
exhibit the data satisfying irrelevant options.

@@

\tname{} leverages public data sets of benign traffic~\cite{tcpreplay}
and additional examples of the attack to guide the search for more
general rules to address the overffiting issue. To that end, \tname{}
\emph{minimizes} the rules produced in the previous step, preserving
the invariant of \emph{only} capturing the positive malicious
traffic. We say that a rule is minimal if discarding any option
results in the capture of negative (\ie{}, benign) traffic.  The
procedure we propose in this step builds on the observation that all
options from the rule produced in the previous step are satisifed by
the traffic produced by the attack. Consequently, any subset of these
options also captures the attack. The rule creation procedure
iteratively discards options from visited rules---derived from the
seed rule---using the negative traffic to guide the search for minimal
rules.

\algdef{SE}[DOWHILE]{Do}{doWhile}{\algorithmicdo}[1]{\algorithmicwhile\ #1}%
\algnewcommand\algorithmicforeach{\textbf{for each}}
\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
\algnewcommand\And{\textbf{and} }%
\setlength{\textfloatsep}{3pt}% Remove \textfloatsep
\begin{algorithm}[t!]
  %\noindent\hspace{-2ex}
  \flushleft{}\textbf{INPUT:} (overconstrained) seed rule $sr$, bening traffic
  $\mathit{bTr}$, malli-\\
  cious traffic $\mathit{mTr}$\\
  %\noindent\hspace{-27.5ex} 
  \textbf{OUTPUT:} Set of minimized rules $\mathcal O$\\
%\vspace{-3ex}
\captionof{algorithm}{Rule creation algorithm}\label{algo1}
\begin{algorithmic}[1]
  \State $\mathcal R \gets \{\mathit{sr}\}$\Comment{initial population of candidate solutions.}\label{line-init}
  \State $\mathcal O \gets \emptyset$
  \Do\label{start-do-while}
      \State $\mathcal R' \gets \emptyset$ \Comment{new iteration. reset.}\label{init-rprime}
      \ForEach {$ro \in \mathcal R $}\Comment{analyze each candidate rule.}\label{start-foreach}
           \For {$i \gets 1$ to $N$}\Comment{mutate each rule $N$ times.}\label{start-for}
               \State $r \gets copy(ro)$\label{copy}
               \State $j\gets \mathit{rand}(0,\mathit{len}(r))$\label{choose-index}
               \State $r[j]\gets 0$\Comment{discard option $j$ in $r$}\label{delete-option}
               \If{$\mathit{isNew(r)} \And\mathit{isPlausible}(r, \mathit{bTr})$} 
                   \State $\mathcal R' \gets \mathcal R' \cup \{r\}$\label{line-progress}
               \EndIf
           \EndFor\label{end-for}
       \EndFor\label{end-foreach}
       \State $\mathcal R \gets \mathcal R'$\Comment{update new generation}\label{define-next-gen}
       \State $\mathcal O \gets \mathcal O \cup \mathcal  R$\Comment{remember all rules}\label{output-accum}
  \doWhile{$\mathcal R\not= \{\}$}\Comment{repeat until reaching a
    fixpoint}\label{end-do-while}\\
  \Return $\mathit{filter(\mathcal O, mTr)}$\label{filtering}
\end{algorithmic}
\end{algorithm}

Algorithm~\ref{algo1} shows the pseudocode for deriving rules from the
seed rule produced at the reverse engineering step. The algorithm
takes as input an overconstrained rule $sr$, produced by the reverse
engineering step, and produces on output a set of alternative rules to
$sr$ that captures every instance of the attack ($\mathit{mTr}$), but
does not capture any message from the benign traffic ($\mathit{bTr}$).

Line~\ref{line-init} initializes the current population of individuals
(\ie{}, rules) $\mathcal{R}$ with the seed rule $sr$. We encode a rule
as a bitvector where each index in the vector represents one option;
the value assigned at a given position indicates whether or not the
option is present in the rule. One iteration of the outer loop
(lines~\ref{start-do-while}-\ref{end-do-while}) processes the current
population of individuals $\mathcal R$ and creates the next generation
$\mathcal R'$. Line~\ref{init-rprime} initializes the population of
individuals for the next generation. The inner loop
(lines~\ref{start-foreach}--\ref{end-foreach}) iterates through each
candidate rule in $\mathcal R$. The algorithm mutates each of these
rules for $N$ times, discarding one option of the rule each
time. Line~\ref{copy} assigns a copy of the bitvector, encoding
$\mathcal R$, to variable $r$. Function $\mathit{len}$ returns the
length of a vector and function $\mathit{rand}$ randomly chooses one
integer value in the range defined by its parameters. So, the
expression $\mathit{rand(0, len(r))}$, appearing at
line~\ref{choose-index}, returns an index of the option in $r$ to
delete at line~\ref{delete-option}. The call to $\mathit{isNew}$
checks if $r$ has not been previously generated through another path
whereas the call to $\mathit{isPlausible}$ checks if the rule only
captures positive traffic, \ie{} no message present in $\mathit{bTr}$
is captured by $r$. If $r$ satisfies both criteria, it is added to
$\mathcal R'$, which denotes the next generation of individuals in the
search. After processing every candidate rule in $\mathcal R$, the
next generation of individuals is defined (line~\ref{define-next-gen})
and the set of discovered rules is updated (at
line~\ref{output-accum}).

The algorithm terminates when no rule in the current generation of
individuals can be further reduced without violating the property that
a candidate solution should only capture positive traffic. This means
that execution of one iteration has not reached the statement declared
at line~\ref{line-progress}, which adds one element to the set
$\mathcal{R'}$. As $\mathcal{R'}$ is initialized empty that set will
remain empty at the end of an iteration of the outer loop.

Note that the algorithm produces no rules in the last iteration. Rules
with a single option are produced in the second-to-last iteration. In
other iterations, rules with more than one option are
produced. Conceptually, the search induces a search tree with the seed
rule $sr$ as the root, single-option rules as child nodes, and rules
with more options as internal nodes. The edges of the tree denote the
ancestor relation between two rules.

It is also worth noting that the algorithm reports every rule in the
tree instead of only those that are surely minimal, \ie{} rules with
one option. The reason is that the the traffic used to check
plausibility is limited by definition---reporting only minimal rules
with respect to that traffic could discard the correct rule. In
summary, this algorithm reports many rules that are plausible by using
the benign traffic to reduce the overconstrained rule produced at the
\reveng\ step. In the Raking step (Section~\ref{sec:ranking}),
\tname{} ranks the rules this algorithm reports by their similarity to
existing trusted rules. Intuitively, rules that do not look like those
trusted rules rank low.

\begin{proposition}
  \vspace{-1ex}  
  The rule generation algorithm reports rules that capture only the
  positive traffic.
\end{proposition}
\begin{proof}
  \vspace{-1.5ex} Every rule created during the search captures the
  attack, provided as example on step one, as every rule contains a
  proper subset of the options from the original rule and all those
  options are satisfied by the traffic produced by the attack. By
  definition of function $\mathit{isPlausible}$, these rules also
  capture \emph{only} the positive traffic. If negative traffic is
  captured, a rule is not added to $\mathcal R'$. Consequently, it
  cannot be an element of $\mathcal O$ and reported on the
  output. Finally, only the rules in $\mathcal O$ that capture all the
  instances of the attack (\ie{}, variants) are reported on output.
  \vspace{-1ex}    
\end{proof}

@@

\textbf{Implementation details.} We implemented function
\emph{isPlausible} by 1)~updating Suricata to only monitor the rule
that is passed as parameter\Mar{isto realmente eh necessario? se sim,
  explique. me pareceu que o que vc. precisa ate aqui jah estaria no
  log do \suri}, 2)~replaying the negative benign traffic (see
Section~\ref{sec:dataset-benign}), and 3)~reading the Suricata logs to
check if any benign message was captured by the rule. It is worth
noting that the benign traffic is stored in \emph{pcap}
files~\cite{pcap} and Suricata can process those files very
efficiently compared to analyzing network messages directly. Instead
of sending traffic to a machine with a deployed server (\eg{}, web
server) that \suri\ would monitor, our setup calls \suri\ directly
through an API. \suri\ is invoked multiple times in the process of
minimizing the rule.  \textbf{Limitations.} It is worth noting that
the performance of \tname{} depends on the quality of inputs. For the
positive malicious traffic, \tname{} only requires one sample (as
input to step one), but can benefit of several samples (see
line~\ref{filtering} of Algorithm~\ref{algo1}). The same applies to
the negative traffic. However, in this case, \tname{} can use a large
variety of public data sets available online (see
Section~\ref{sec:objectanalysis-benign}).
